---
title: "Replication of birdwatch paper"
author: "Drishya Shrestha & Naomi Beck"
date: "`r Sys.Date()`"
output: html_document
---
 
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(scales)
library(modelr)
library(readr)
library(ggplot2)

theme_set(theme_bw())
```
 
 
```{r load_data}
setwd("./community-notes-2025-group-1")
ratings <- read_tsv("ratings_jan23_to_july31.tsv")
str(ratings)
view(head(ratings))

notes <- read_tsv("notes_jan23_to_july31.tsv")
view(head(notes))
summary(notes)
str(notes)
 
```

# Figure 2: Number of users who responded “Yes” to the question “Did you link to sources you believe most people would consider trustworthy?”
```{r figure_2, echo=FALSE}

notes %>% 
 mutate(
    trustworthySources = factor(
      trustworthySources,
      levels = c(0, 1), #actual values
      labels = c("Not Trustworthy", "Trustworthy")
    ),
    classification = factor(
      classification,
      levels = c("NOT_MISLEADING", "MISINFORMED_OR_POTENTIALLY_MISLEADING"), # aactual values
      labels = c("Not Misleading", "Misleading")
    ))%>%
    group_by(classification,trustworthySources) %>% 
    summarise(total_num = n(), .groups ="drop") %>% 
        ggplot(aes(total_num,classification, fill= trustworthySources))+
        geom_bar(stat="identity")+
        labs(title = "Notes by Classification and Trustworthy Sources",
       y = "Classification",
       x = "Number of Notes"
       ) +
       scale_fill_manual(values = c("Not Trustworthy" = "#ffc70e", "Trustworthy" = "#1b58e7"))

```


# Figure 3: Number of Birdwatch notes per checkbox answer option in response to the question “Why do you believe this tweet may be misleading?”
```{r figure_3, echo=FALSE}
library(ggplot2)

# Select relevant columns
misleading_cols <- c(
  "misleadingFactualError",
  "misleadingMissingImportantContext",
  "misleadingUnverifiedClaimAsFact",
  "misleadingOutdatedInformation",
  "misleadingOther",
  "misleadingSatire",
  "misleadingManipulatedMedia"
)

# Tally up the 1s
misleading_counts <- colSums(notes[misleading_cols] == 1, na.rm = TRUE)
misleading_counts

# Create a data frame for plotting
misleading_counts_df <- data.frame(
  Category = c(
    "Factual error",
    "Missing important context",
    "Unverified claim as fact",
    "Outdated information",
    "Other",
    "Satire",
    "Manipulated media"
  ),
  Count = misleading_counts
)

# Plot
ggplot(misleading_counts_df, aes(x = Count, y = reorder(Category, Count))) +
  geom_bar(stat = "identity", fill = "firebrick") +
  labs(
    x = "Number of Birdwatch Notes",
    y = NULL
  ) 
```


# Figure4: Number of Birdwatch notes per check box answer optionin response to the question “Why do you believe this tweet is not misleading?”
```{r figure_4, echo=FALSE}

# Select relevant columns
not_misleading_cols <- c(
  "notMisleadingFactuallyCorrect",
  "notMisleadingPersonalOpinion",
  "notMisleadingClearlySatire",
  "notMisleadingOther",
  "notMisleadingOutdatedButNotWhenWritten"
)

# Tally up the 1s
not_misleading_counts <- colSums(notes[not_misleading_cols] == 1, na.rm = TRUE)

# Create a data frame for plotting
not_misleading_counts_df <- data.frame(
  Category = c(
    "Factually correct",
    "Personal opinion",
    "Clearly satire",
    "Other",
    "Outdated but not wen written"
  ),
  Count = not_misleading_counts
)

# Plot
ggplot(not_misleading_counts_df, aes(x = Count, y = reorder(Category, Count))) +
  geom_bar(stat = "identity", fill = "navy") +
  labs(
    x = "Number of Birdwatch Notes",
    y = NULL
  ) 
```



#  Figure 5: CCDFs for word count in text explanations of Birdwatch notes.
```{r figure_5c, echo=FALSE}



graph_5c <- notes %>% mutate(
    classification = factor(
      classification,
      levels = c("NOT_MISLEADING", "MISINFORMED_OR_POTENTIALLY_MISLEADING"), # aactual values
      labels = c("Not Misleading", "Misleading")
)) %>%
    #counting the url as one word
    mutate(word_count = str_count(summary, "\\w+"))%>%
    view(head(10))

summary(graph_5c)
graph_5c

filter(graph_5c, word_count == max(word_count, na.rm = TRUE)) %>%
  select(summary, word_count) %>% view()

    
##GRAPH
notes %>%
  mutate(
    classification = factor(
      classification, 
      levels = c("NOT_MISLEADING", "MISINFORMED_OR_POTENTIALLY_MISLEADING"),
      labels = c("Not Misleading", "Misleading")
    ),
    summary_clean = str_replace_all(summary, "https?://[^\\s]+", "URL"),
    word_count = str_count(summary_clean, "\\w+")
  ) %>%
    group_by(classification, word_count) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(classification) %>%
    arrange(word_count) %>% 
    mutate(ccdf = 1 - cumsum(count)/sum(count))%>% 
    ggplot(aes(word_count,ccdf, color = classification))+
    geom_line()  + scale_y_continuous(labels = percent)+
    scale_y_log10()+
    scale_color_manual(
    values = c("Not Misleading" = "blue", "Misleading" = "red")
  )


filtered_graph_5c <- notes %>%
  mutate(
    classification = factor(
      classification, 
      levels = c("NOT_MISLEADING", "MISINFORMED_OR_POTENTIALLY_MISLEADING"),
      labels = c("Not Misleading", "Misleading")
    ),
    summary_clean = str_replace_all(summary, "https?://[^\\s]+", "URL"),
    word_count = str_count(summary_clean, "\\w+")
  )
    
summary(filtered_graph_5c)





```


# Figure_7_CCDFs for (a)helpfulness ratio and (b)total votes.
```{r figure_7a, echo=FALSE}
notes_clean <- notes %>% mutate(noteId = as.character(noteId))
notes_ratings <- left_join(notes_clean, ratings, by = "noteId")
view(head(notes_ratings,100))
view(head(ratings, 100))
notes_ratings %>%  
    mutate(
    classification = factor(
      classification, 
      levels = c("NOT_MISLEADING", "MISINFORMED_OR_POTENTIALLY_MISLEADING"),
      labels = c("Not Misleading", "Misleading")
    ),
    #To account for both versions creating a new column which tells if its helpful or not 
   helpful_num = as.numeric(helpfulnessLevel %in% c("HELPFUL", "SOMEWHAT_HELPFUL") | helpful %in% c("1")))%>% 
   select(noteId,helpful_num, classification) %>%
   #

    group_by(noteId, classification)%>%
    summarise(helpfulness_ratio = mean(helpful_num, na.rm=TRUE),.groups = "drop")%>%

    group_by(classification)%>%
    arrange(helpfulness_ratio)%>%
    mutate(ccdf = 1 - cumsum(helpfulness_ratio)/sum(helpfulness_ratio))%>%

    ggplot(aes(helpfulness_ratio, ccdf, color = classification))+
    geom_line() + scale_y_continuous(labels = percent)+
    scale_color_manual(values = c("Not Misleading" = "blue", "Misleading" = "red")
)

```


```{r figure_7b, echo=FALSE}
notes_ratings %>%  
    mutate(
    classification = factor(
      classification, 
      levels = c("NOT_MISLEADING", "MISINFORMED_OR_POTENTIALLY_MISLEADING"),
      labels = c("Not Misleading", "Misleading")
    ),
    helpful = as.numeric(helpful),
    notHelpful = as.numeric(notHelpful))%>%
    group_by(noteId,classification)%>%
    summarise(total_votes = sum(helpful, na.rm = TRUE)+sum(notHelpful, na.rm = TRUE), .groups = "drop") %>%
    group_by(classification, total_votes) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(classification)%>%
    arrange(total_votes)%>%
    mutate(ccdf = 1 - (row_number() - 1) / n())%>%
    ggplot(aes(total_votes, ccdf, color = classification))+
    geom_line() + scale_y_continuous(labels = percent) +
    scale_color_manual(values = c("Not Misleading" = "blue", "Misleading" = "red")
)
```



# Figure8: Number of ratings per check box answer optionin response to the prompt “What about this note was helpful to you?”
```{r figure_8, echo=FALSE}
# Select relevant columns
helpful_cols <- c(
  "helpfulInformative",
  "helpfulClear",
  "helpfulGoodSources",
  "helpfulEmpathetic",
  "helpfulUniqueContext",
  "helpfulAddressesClaim",
  "helpfulImportantContext",
  "helpfulOther"
)

# Tally up the 1s
helpful_counts <- colSums(ratings[helpful_cols] == 1, na.rm = TRUE)

# Create a data frame for plotting
helpful_counts_df <- data.frame(
  Category = c(
    "Informative",
    "Clear",
    "Good sources",
    "Empathetic",
    "Unique context",
    "Addresses claim",
    "Important context",
    "Other"
  ),
  Count = helpful_counts
)

# Plot
ggplot(helpful_counts_df, aes(x = Count, y = reorder(Category, Count))) +
  geom_bar(stat = "identity", fill = "navy") +
  labs(
    x = "Number of Ratings",
    y = NULL
  ) 

```


# Figure 9: Number of ratings per check box answer option in response to the prompt “Help us understand why this note was not helpful to you?”
```{r figure_9, echo=FALSE}
# Select relevant columns
not_helpful_cols <- c(
    "notHelpfulSourcesMissingOrUnreliable",
    "notHelpfulOpinionSpeculationOrBias",
    "notHelpfulMissingKeyPoints",
    "notHelpfulArgumentativeOrBiased",  #inflammatory
    "notHelpfulIncorrect",
    "notHelpfulOffTopic",
    "notHelpfulOther",
    "notHelpfulHardToUnderstand",
    "notHelpfulSpamHarassmentOrAbuse",
    "notHelpfulOutdated", 
    "notHelpfulIrrelevantSources"
)

# Tally up the 1s
not_helpful_counts <- colSums(ratings[not_helpful_cols] == 1, na.rm = TRUE)

# Create a data frame for plotting
not_helpful_counts_df <- data.frame(
  Category = c(
    "Sources missing or unreliable",
    "Opinion speculation or bias",
    "Missing key points",
    "Argumentative or inflammatory",
    "Incorrect",
    "Off topic",
    "Other",
    "Hard to understand",
    "Spam harassment or abuse",
    "Outdated",
    "Irrelevant sources"
  ),
  Count = not_helpful_counts
)

# Plot
ggplot(not_helpful_counts_df, aes(x = Count, y = reorder(Category, Count))) +
  geom_bar(stat = "identity", fill = "firebrick") +
  labs(
    x = "Number of Ratings",
    y = NULL
  ) 
```
